---
title: "Status_setup_diabetes"
author: "NN"
date: "2024-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Censoring data by type 1 diabetes, death, immigration (lost to follow-up), or not experiencing event (diabetes) before end of follow-up.

Not censored = 0, Censored due to T1D or Death = 1, Event (T2D) = 2. 

All individuals with codes 20, 30, 50 and 60 in the CPR registry, denoting somehow faulty entries, to be removed from the dataset.

The df dod2020 is incomplete compared to the t_person file in terms of date of death (6063 NA's), thus will just use t_person for info on date of death.

In terms of cause of death the number of entries do not correspond with the number of individuals who have died. File dodaarsaag_1 spans 1970-2001, while file _2 spans 2002-2018 
```{r}

load("E:/workdata/707957/_DataIn/R/Basic/kronikere.rdata")
load("E:/workdata/707957/_DataIn/R/Basic/t_person.rdata")

#Coding status variable + outdate for event T2D
kronikere$status[kronikere$sygdom == "TYPE 2"] <- 2
kronikere$status[kronikere$sygdom == "TYPE 1"] <- 3
kronikere <- kronikere %>% 
  rename(outdate = debut) %>% 
   select(pnr,
         status,
         outdate)
#A total of 17590 individuals with T1D, 38607 with T2D.

#Coding censoring variable from death/emigration
t_person <- t_person %>%
  filter(C_STATUS < 10 | C_STATUS >= 70) %>% 
  mutate(status = ifelse(C_STATUS == 90, 1,0)) %>% 
  rename(outdate = D_STATUS_HEN_START,
         pnr = v_pnr) %>% 
  select(pnr,
         status,
         outdate)

#55978 appear to have died, while 2906949 have survived (or emigrated)
  
#Gathering into status/outage df 
dfstat <- rbind(t_person, kronikere) %>%
  group_by(pnr) %>% 
  slice(which.max(status))

#Making all outdate = NA into 31.12.2018
eof <- as.Date("2018-12-31")
dfstat$outdate[is.na(dfstat$outdate)] <- eof

rm(kronikere, t_person)

save(dfstat, file="E:\\workdata\\707957\\ID-NN\\Results\\status.rdata")
