---
title: "t2d_setup"
author: "NN"
date: "2024-01-08"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
```
## Prepare data

```{r }
#load data
load("E:/workdata/707957/_DataIn/R/Basic/t_scor_1972_1987.rdata")
load("E:/workdata/707957/_DataIn/R/Basic/t_dodsaarsag_1.rdata")
load("E:/workdata/707957/_DataIn/R/Basic/t_dodsaarsag_2.rdata")
load("E:/workdata/707957/_DataIn/R/Basic/t_person.rdata")
load("E:/workdata/707957/_DataIn/R/Basic/samlpop_707957.rdata")

#Renaming pnr variable
t_scor_1972_1987 <- rename(t_scor_1972_1987, pnr = cprnr)
t_dodsaarsag_1 <- rename(t_dodsaarsag_1, pnr = k_cpr)
t_dodsaarsag_2 <- rename(t_dodsaarsag_2, pnr = k_cpr)
t_person <- rename(t_person, pnr = v_pnr)

#Merging basic SCOR datasets
scor <- left_join(t_scor_1972_1987, samlpop_707957, by="pnr") %>%
  distinct() # Removing all duplicates

#Creating variable for year of birth and year of visit
scor$visit_year <- format(as.Date(scor$statusdato, format="%d/%m/%Y"),"%Y")
scor$birth_year <- format(as.Date(scor$fodedto, format="%d/%m/%Y"),"%Y")


#Calculating age at death
scor <- left_join(scor, t_dodsaarsag_1, by="pnr") %>% left_join(., t_dodsaarsag_2, by = "pnr")
require(lubridate)
scor$age_at_death <-trunc((scor$fodedto %--%scor$D_DODSDTO)/years(1))

#Calculating age at examination
scor$ageatexam <-trunc((scor$fodedto %--%scor$statusdato)/years(1))

#Create grouping into valid and invalid entries (dead before 1972, age at examination above 19 (N invalid entries=981)
scor$validage[scor$ageatexam >= 0] <- "Valid"
scor$validage[scor$ageatexam >= 19]<- "Invalid"
scor$validage[scor$age_at_death <= 1] <- "Invalid"
scor$validage[scor$currentage<34] <- "Invalid"

#Removing invalid entries
scor <- scor[(scor$validage =="Valid"),]

#Recoding PATOLOGISKE_POCHER for 0/1 instead of NA/X
scor$PATOLOGISKE_POCHER[scor$PATOLOGISKE_POCHER == "NA"] <- 0
scor$PATOLOGISKE_POCHER[scor$PATOLOGISKE_POCHER == "X"] <- 1
scor$perio <- as.numeric(scor$PATOLOGISKE_POCHER)

#Make gingivitis index numeric
scor$gingivitis <- as.numeric(scor$gingivitisindex)


scor <- scor %>%
  left_join(., t_person, by = "pnr") %>%  
  filter(C_STATUS < 10 | C_STATUS >= 70) %>% #n=656092 when removing invalid CPR-numbers after filtering for birth years of interest.
  select(pnr,
         statusdato,
         koen,
         fodedto,
         birth_year,
         visit_year,
         ageatexam,
         gingivitis,
         perio) %>%
  filter(birth_year>=1963 & birth_year<=1972) #filter variables and birth years of interest n = #n= 657447


rm(samlpop_707957, t_dodsaarsag_1, t_dodsaarsag_2, t_person)

load("E:/workdata/707957/ID-NN/Results/dmf.rdata")
load("E:/workdata/707957/ID-NN/Results/status.rdata")

#Load, status, and dmf files
scor <- scor %>%
  inner_join(., dmf, by=c("pnr", "statusdato")) %>%
  inner_join(., dfstat, by = "pnr") #656092 individuals

#inage/outage calculation

scor$status[scor$outdate >"2018-12-31"] <- "0" #If event is after end of follow-up in 2018 set outdate to end of follow-up and status to 0
scor$outdate[scor$outdate >"2018-12-31"] <- "2018-12-31"

scor <- scor %>%
  mutate(fodedto = as.Date(fodedto, format = "%Y/%m/%d")) %>%
  mutate(outdate = as.Date(outdate, format = "%Y/%m/%d")) #just to make sure all ages calculate on the same basis.

require(lubridate)
scor$outage <-((scor$fodedto %--%scor$outdate)/years(1)) #Calculating age at event/censoring

sof <- as.Date("1995-01-01") #Making start of follow-up date - baseline 1995
scor$indate <- sof
 
scor$inage <-((scor$fodedto %--%scor$indate)/years(1)) #Calculating age at start of follow-up in 1995

subset <- scor %>%
  filter(inage <25) %>%
  mutate(inage = 25) #filter for individuals young than 25 in 1995, set start of follow-up to age 25 for these

scor <- scor %>%
  filter(inage >=25) %>%
  rbind(., subset) #merge with subset. Total number of individuals in dataset 650,092
  
scor <- scor %>%
  filter(outdate >= "1995-01-01" & outage >=25) #Filter to remove individuals who have event or are censored before beginning of follow-up on the first of January 1995 OR before age 25, n=635,031/ 21,061 individuals removed ,

dif <- scor %>% filter(outage < inage) #removing those who have event before enrollment/age 25 (none are filtered out, good).

#Time on study variable for descriptive stats
scor$tos <- scor$outage-scor$inage

#Making binary and categorical variables for gingivitis and DMFT
scor <- scor %>%
    mutate(gingi.bi = ifelse(gingivitis > 0, 1, 0),
          dmft.bi = ifelse(dmft > 0, 1, 0))

#Categorical gingi and dmft - updated categorisation based on model checks

scor$gingi.cat[scor$gingivitis >= 0 & scor$gingivitis <=4] <- 0
scor$gingi.cat[scor$gingivitis >= 5 & scor$gingivitis <=8] <- 1
scor$gingi.cat[scor$gingivitis >=9 & scor$gingivitis <=12] <- 2
 

scor$dmft.cat[scor$dmft >= 0 & scor$dmft <=4] <- 0
scor$dmft.cat[scor$dmft >= 5 & scor$dmft <=12] <- 1
scor$dmft.cat[scor$dmft >=13 & scor$dmft <=16] <- 2
scor$dmft.cat[scor$dmft >=17] <- 3

#Making combined caries-gingi variable
 
scor$comb[scor$dmft.cat == 0 & scor$gingi.cat == 0] <- 0
scor$comb[scor$dmft.cat == 0 & scor$gingi.cat == 1] <- 1
scor$comb[scor$dmft.cat == 0 & scor$gingi.cat == 2] <- 2
scor$comb[scor$dmft.cat == 1 & scor$gingi.cat == 0] <- 3
scor$comb[scor$dmft.cat == 1 & scor$gingi.cat == 1] <- 4
scor$comb[scor$dmft.cat == 1 & scor$gingi.cat == 2] <- 5
scor$comb[scor$dmft.cat == 2 & scor$gingi.cat == 0] <- 6
scor$comb[scor$dmft.cat == 2 & scor$gingi.cat == 1] <- 7
scor$comb[scor$dmft.cat == 2 & scor$gingi.cat == 2] <- 8
scor$comb[scor$dmft.cat == 3 & scor$gingi.cat == 0] <- 9
scor$comb[scor$dmft.cat == 3 & scor$gingi.cat == 1] <- 10
scor$comb[scor$dmft.cat == 3 & scor$gingi.cat == 2] <- 11


save(scor, file="E:\\workdata\\707957\\ID-NN\\Results\\t2d_df.rdata")
```

#Severity over time dataset prep

```{r}

load("E:/workdata/707957/ID-NN/Results/t2d_df.rdata")

#Below age 6
df1 <- scor %>%
  select(pnr,
         dmft,
         gingivitis,
         ageatexam) %>%
  filter(ageatexam <=5.999)

df <- df1 %>%
  select(!gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(dmft))
  
df1 <- df1 %>%
  select(pnr, gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(gingivitis)) %>%
   left_join(., df,  by="pnr") %>%
  rename(caries1 = dmft,
         gingi1 = gingivitis)
  
#Age 6 to 12
df2 <- scor %>%
  select(pnr,
         dmft,
         gingivitis,
         ageatexam) %>%
  filter(ageatexam >=6.000 & ageatexam <= 11.999)

df <- df2 %>%
  select(!gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(dmft))
  
df2 <- df2 %>%
  select(pnr, gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(gingivitis)) %>%
   left_join(., df,  by="pnr") %>%
    rename(caries2 = dmft,
         gingi2 = gingivitis)

#Age 12 and up
df3 <- scor %>%
  select(pnr,
         dmft,
         gingivitis,
         ageatexam) %>%
  filter(ageatexam >=12.000)

df <- df3 %>%
  select(!gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(dmft))
  
df3 <- df3 %>%
  select(pnr, gingivitis) %>%
  group_by(pnr) %>%
  slice(which.max(gingivitis)) %>%
   left_join(., df,  by="pnr") %>%
    rename(caries3 = dmft,
         gingi3 = gingivitis)

df <- full_join(df1, df2, by ="pnr") %>%
  full_join(., df3, by = "pnr") %>%
  select(!c(ageatexam.x, ageatexam.y, ageatexam))

df <- df %>% na.omit()

df <- df %>%
  group_by(pnr) %>%
  mutate(cariesload = sum(caries1+caries2+caries3),
         gingiload = sum(gingi1+gingi2+gingi3),
         caries_time1 = ifelse(caries1 > 0, 1, 0),
         caries_time2 = ifelse(caries2 > 0, 1, 0),
         caries_time3 = ifelse(caries3 > 0, 1, 0),
         gingi_time1 = ifelse(gingi1 > 0, 1, 0),
         gingi_time2 = ifelse(gingi2 > 0, 1, 0),
         gingi_time3 = ifelse(gingi3 > 0, 1, 0),
         caries_duration = sum(caries_time1+caries_time2+caries_time3),
         gingi_duration = sum(gingi_time1+gingi_time2+gingi_time3))

#Grouping gingi duration

df$gingi_duration[df$gingi_duration <=1] <- 0
df$gingi_duration[df$gingi_duration ==2] <- 1
df$gingi_duration[df$gingi_duration ==3] <- 2

df$caries_duration[df$caries_duration <=1] <- 0
df$caries_duration[df$caries_duration ==2] <- 1
df$caries_duration[df$caries_duration ==3] <- 2

#grouping gingiload: Aim: which group an individuals belongs to determined by highest score at any one visit being within the boundaries of the original GS grouping so:
#Group 0 -> no total score higher than 12 (i.e. 3*4), no single score greater than 4. Group 1 -> total score greater than 5 (meaning no single score less than 4 or greater than 8 (so at least one score of a moderate severity needed).
#Group 3 -> At least one registration of 9 (i.e. registration of high level of gingivitis at all registrations needed).

df$ggroup[df$gingi1 <= 4  & df$gingi2 <= 4 & df$gingi3 <= 4] <- 0
df$ggroup[df$gingi1 >=5 & df$gingi1 <=8 | df$gingi2 >=5 & df$gingi2 <=8 | df$gingi3 >=5 & df$gingi3 <=8] <- 1
df$ggroup[df$gingiload >=9 & df$gingi1 >=9|df$gingi2 >=9|df$gingi3 >=9] <- 2
#distribution reflects the normal gingi.cat variable well (stratified on male and female).


#grouping cariesload. Like with gingivitis above aim to group individuals by the upper and lower limits of their registrations. Group 0: No single registration higher than 4. Group 1: No single registration greater than 16, BUT total cariesload no greater than 24, meaning that an individual can't have more no more than


df$cgroup[df$caries1 <=4  & df$caries2 <=4 & df$caries3 <=4] <- 0
df$cgroup[df$caries1 >=5 & df$caries1 <= 12 |df$caries2 >=5 & df$caries2 <= 12 | df$caries3 >=5 & df$caries3 <= 12] <- 1
df$cgroup[df$caries1 >= 13 |df$caries2 >= 13 |df$caries2 >= 13] <- 2
#df$cgroup[df$cariesload >=36] <- 3
#distribution seems to fit pretty well with dmft.cat variable

save(df, file="E:\\workdata\\707957\\ID-NN\\Results\\timevary.rdata")

rm(df, df1, df2, df3)
```
## Set-up for analyses with highest achieved level of education between ages 25-30

```{r}

load("E:/workdata/707957/ID-NN/Results/t2d_df.rdata")
load("E:/workdata/707957/ID-NN/Results/status.rdata")

scor <- scor %>%
  select(!c(status, outdate)) %>%
  left_join(., dfstat, by ="pnr")

#Select on highest ever registered level of gingivitis
gingi <- scor %>%
  select(pnr, gingivitis, gingi.cat, gingi.bi, ageatexam) %>%
  group_by(pnr) %>%
  slice(which.max(gingivitis)) %>%
  rename(ageatgingi = ageatexam)

#Select on highest registered dmft and join with gingi
scor <- scor %>%
  select(!c(gingivitis, gingi.cat, gingi.bi)) %>%
  rename(ageatdmft = ageatexam) %>%
  group_by(pnr) %>%
  slice(which.max(dmft)) %>%
  left_join(., gingi,  by="pnr")

#636,231 individuals in dataset
rm(gingi, dfstat)

load("E:/workdata/707957/ID-NN/Results/full_edu.rdata")

edu <- edu %>%
  select(pnr, edustat, eduage) %>%
  filter(eduage >= 25 & eduage <=30) %>% #select for education level between ages 25 and 30
  group_by(pnr) %>%
  slice(which.max(edustat))

#merge with SCOR data
test <- left_join(scor, edu, by = "pnr")

scor <- test %>%
  subset(!(is.na(edustat)))

dif <- subset(test, !pnr %in% scor$pnr) #7273 individuals removed due to missing edu.Total n=627,758.

subset <- scor %>%
  filter(inage == outage) %>%
  mutate(outage = outage+0.001)
scor <- scor %>%
  filter(inage < outage) %>%
  rbind(., subset) #for the 31 individuals who both enter and exit follow-up on 01-01-1995 add 0.0001 to avoid issue with inage and outage being identical.

save(scor, file="E:\\workdata\\707957\\ID-NN\\Results\\t2d_df1.rdata")
save(dif, file="E:\\workdata\\707957\\ID-NN\\Results\\leftout.rdata")

```

#Set-up for analysis of subset

```{r}
load("E:/workdata/707957/ID-NN/Results/timevary.rdata")
load("E:/workdata/707957/ID-NN/Results/t2d_df1.rdata")

scor <- scor %>%
  inner_join(., df, by="pnr") %>%
    select(pnr,
           koen,
           ageatgingi,
           ageatdmft,
           inage,
           outage,
           status,
           edustat,
           gingiload,
           cariesload,
           caries_duration,
           gingi_duration,
           ageatgingi,
           ageatdmft,
           cgroup,
           ggroup,
           dmft.cat,
           gingi.cat,
           birth_year)


#Grouping age at dmft and gingi based on variable assessment.

#scor$ageatdmft[scor$ageatdmft <=6] <- 0
#scor$ageatdmft[scor$ageatdmft >=6 & scor$ageatdmft <12] <- 1
scor$ageatdmft[scor$ageatdmft <12] <- 0
scor$ageatdmft[scor$ageatdmft >=12] <- 1


#scor$ageatgingi[scor$ageatgingi <=6] <- 0
#scor$ageatgingi[scor$ageatgingi >=6 & scor$ageatgingi <12] <- 1
scor$ageatgingi[scor$ageatgingi <12] <- 0
scor$ageatgingi[scor$ageatgingi >=12] <- 1

#Grouping age at dmft and gingi based on variable assesment

#scor$gingi_duration[scor$gingi_duration <=1] <- 0
scor$gingi_duration[scor$gingi_duration <2] <- 0
scor$gingi_duration[scor$gingi_duration ==2] <- 1

#scor$caries_duration[scor$caries_duration <=1] <- 0
scor$caries_duration[scor$caries_duration <2] <- 0
scor$caries_duration[scor$caries_duration ==2] <- 1

rm(df)
```

#Alternative comb variable

```{r}
scor$comb[scor$dmft.cat == 0 & scor$gingi.cat == 0] <- 0
scor$comb[scor$dmft.cat == 0 & scor$gingi.cat >= 1] <- 1
scor$comb[scor$dmft.cat == 1 & scor$gingi.cat == 0] <- 2
scor$comb[scor$dmft.cat == 1 & scor$gingi.cat >= 1] <- 3
scor$comb[scor$dmft.cat >= 2 & scor$gingi.cat == 0] <- 4
scor$comb[scor$dmft.cat >= 2 & scor$gingi.cat >= 1] <- 5
 
```
#Tested but not included

## Set-up for analyses with edu as time-varying variable

```{r}

load("E:/workdata/707957/ID-NN/Results/t2d_df.rdata")

#Select on highest ever registered level of gingivitis
gingi <- scor %>%
  select(pnr, gingivitis, ageatexam) %>%
  group_by(pnr) %>%
  slice(which.max(gingivitis)) %>%
  rename(ageatgingi = ageatexam)

#Select on highest registered dmft and join with gingi
scor <- scor %>%
  select(!c(gingivitis)) %>%
  rename(ageatdmft = ageatexam) %>%
  group_by(pnr) %>%
  slice(which.max(dmft)) %>%
  left_join(., gingi,  by="pnr")

#Using tmerge for time varying covariate Edu

load("E:/workdata/707957/ID-NN/Results/full_edu.rdata")

edu20 <- edu %>%
  select(pnr, edustat, eduage) %>%
  filter(eduage >= 18 & eduage <= 22) %>% #select for education level for age 20
  group_by(pnr) %>%
  slice(which.max(edustat))

test <- edu %>%
  select(pnr, edustat, eduage) %>%
  filter(eduage > 22 & eduage <= 30) %>% #select for education level between ages 20 and 30
  group_by(pnr) %>%
  slice(which.max(edustat)) %>%
  rbind(., edu20) %>%
  mutate(eduage = if_else(eduage > 22, 30, 22))


df1 <- tmerge(scor, scor, id=pnr,  tstart=inage, tstop=outage, status=event(outage, status)) #set range
scor_temp <- tmerge(df1, test, id=pnr, edu = tdc(eduage, edustat))#add time varying variable for education

scor2 <- scor_temp %>%
  group_by(pnr) %>%
  subset(!(is.na(edu)))

dif2 <- subset(scor_temp, !pnr %in% scor2$pnr) #Filtering out those removed from primary dataset for comparative analysis. 4532 individuals removed due to missing info on education, 631,699 individuals left in datastet.Difference between the two approaches to edu = 68 observations/individuals, no T2D cases. Removed due to missing education at age 20.


save(scor2, file="E:\\workdata\\707957\\ID-NN\\Results\\t2d_df2.rdata")
save(dif2, file="E:\\workdata\\707957\\ID-NN\\Results\\leftout2.rdata")
