---
title: "t2d_supplementary"
author: "NN"
date: "2024-11-25"
output: html_document
---

```{r setup, include=FALSE}
#load packages
library(tidyverse)
library(prodlim)
library(survival)
library(survminer)
library(timereg)
library(mets)

#load dataset
load("E:/workdata/707957/ID-NN/Results/JoDR_df1.rdata")

#add 1 to all dmft and gingi to avoid 0 for form assessment to work
scor <- scor %>% 
  group_by(pnr) %>% 
  mutate(dmft = sum(dmft+1),
         gingi = sum(gingivitis+1))
 
females <- scor %>% 
  filter(koen==2)

males <- scor %>% 
  filter(koen==1)
```

## Assessing functional form and proportional hazards of continuous variables

```{r}

#Females
ggcoxfunctional(Surv(inage, outage, status==2)~ dmft + dmft+log(dmft)+sqrt(dmft), data = females) #regular DMFT looks pretty close to linear, go with that.

ggcoxfunctional(Surv(inage, outage, status==2)~ gingi+log(gingi)+sqrt(gingi), data = females) # not quite as linear as dmft, but would still say pretty decent if using regular gingi

#Males
ggcoxfunctional(Surv(inage, outage, status==2)~ dmft + dmft+log(dmft)+sqrt(dmft), data = males) #again regular DMFT looks gives the best linear fit.

ggcoxfunctional(Surv(inage, outage, status==2)~ gingi + log(gingi)+sqrt(gingi), data = males) # pretty good linear fit for regular gingi

#Proportional hazards

fit <- coxph(Surv(inage, outage, status==2)~dmft+gingi+edustat, data=females)
test.ph <- cox.zph(fit)
ggcoxzph(test.ph) #don't see any (strong) systematic departures from the smoothing spline, distribution across time pretty even (thinning at the end)

fit <- coxph(Surv(inage, outage, status==2)~dmft+gingi+edustat, data=males)
test.ph <- cox.zph(fit)
ggcoxzph(test.ph) #again don't see any strong systematic departures from the smoothing spline, distribution across time pretty even (thinning at the end)

#Proportional hazards for dmft.cat og gingicat revisited
#MALES
out=survfit(Surv(inage, outage, status==2)~gingi.cat, data=males)
kmplot(out, fun="cloglog", lty=1) #again curves certainly parallel from age 30

out=survfit(Surv(inage, outage, status==2)~dmft.cat, data=males)
kmplot(out, fun="cloglog", lty=1) #curves parallel from age 40

#FEMALES
out=survfit(Surv(inage, outage, status==2)~gingi.cat, data=females)
kmplot(out, fun="cloglog", lty=1) #would say curves are parallel from age 30


out=survfit(Surv(inage, outage, status==2)~dmft.cat, data=females)
kmplot(out, fun="cloglog", lty=1) #again curves certainly parallel from age 40
```

## Running models for continuous variables

```{r}
summary(coxph(Surv(inage, outage, status==2)~dmft+strata(edustat), data=females))
         coef exp(coef) se(coef)    z Pr(>|z|)    
dmft 0.016187  1.016318 0.002781 5.82 5.87e-09 ***
     exp(coef) exp(-coef) lower .95 upper .95
dmft     1.016     0.9839     1.011     1.022

summary(coxph(Surv(inage, outage, status==2)~gingi+strata(edustat), data=females))
          coef exp(coef) se(coef)     z Pr(>|z|)    
gingi 0.075794  1.078740 0.005034 15.06   <2e-16 ***
      exp(coef) exp(-coef) lower .95 upper .95
gingi     1.079      0.927     1.068     1.089

summary(coxph(Surv(inage, outage, status==2)~dmft + gingi, data=females))
          coef exp(coef) se(coef)      z Pr(>|z|)    
dmft  0.019526  1.019718 0.002849  6.852 7.26e-12 ***
gingi 0.092926  1.097381 0.005224 17.787  < 2e-16 ***
      exp(coef) exp(-coef) lower .95 upper .95
dmft      1.020     0.9807     1.014     1.025
gingi     1.097     0.9113     1.086     1.109

summary(coxph(Surv(inage, outage, status==2)~dmft + gingi+strata(edustat), data=females))
          coef exp(coef) se(coef)      z Pr(>|z|)    
          coef exp(coef) se(coef)      z Pr(>|z|)    
dmft  0.006004  1.006022 0.002869  2.093   0.0364 *  
gingi 0.073081  1.075818 0.005199 14.058   <2e-16 ***
      exp(coef) exp(-coef) lower .95 upper .95
dmft      1.006     0.9940     1.000     1.012
gingi     1.076     0.9295     1.065     1.087

summary(coxph(Surv(inage, outage, status==2)~dmft +gingi+dmft*gingi+strata(edustat), data=females)) #no interactions found

summary(coxph(Surv(inage, outage, status==2)~dmft+strata(edustat), data=males))
         coef exp(coef) se(coef)     z Pr(>|z|)    
dmft 0.010594  1.010651 0.002206 4.801 1.58e-06 ***
     exp(coef) exp(-coef) lower .95 upper .95
dmft     1.011     0.9895     1.006     1.015

summary(coxph(Surv(inage, outage, status==2)~gingi+strata(edustat), data=males))
          coef exp(coef) se(coef)     z Pr(>|z|)    
gingi 0.065503  1.067696 0.004002 16.37   <2e-16 ***
      exp(coef) exp(-coef) lower .95 upper .95
gingi     1.068     0.9366     1.059     1.076

summary(coxph(Surv(inage, outage, status==2)~dmft + gingi, data=males))
           coef exp(coef) se(coef)      z Pr(>|z|)    
dmft  0.013489  1.013580 0.002257  5.976 2.29e-09 ***
gingi 0.081551  1.084969 0.004147 19.666  < 2e-16 ***
          exp(coef) exp(-coef) lower .95 upper .95
dmft      1.014     0.9866     1.009     1.018
gingi     1.085     0.9217     1.076     1.094

summary(coxph(Surv(inage, outage, status==2)~dmft + gingi+strata(edustat), data=males))
          coef exp(coef) se(coef)      z Pr(>|z|)    
dmft  0.001638  1.001640 0.002274  0.721    0.471    
gingi 0.064769  1.066913 0.004130 15.682   <2e-16 ***
      exp(coef) exp(-coef) lower .95 upper .95
dmft      1.002     0.9984    0.9972     1.006
gingi     1.067     0.9373    1.0583     1.076

summary(coxph(Surv(inage, outage, status==2)~dmft +gingi+dmft*gingi+strata(edustat), data=males)) #no interactions found

```

# Cumulative hazard plots 

```{r}

out=survfit(Surv(inage, outage, status==2)~dmft.cat, data=females)
ggsurvplot(out, fun = "cumhaz", xlim=c(25,56), size=0.05, conf.int="TRUE", legend.labs=c("DCS: Low", "DCS: Moderate", "DCS: High", "DCS: Severe"), xlab ="Age during follow-up", ylab="Cumulative hazard of T2D in females")

out=survfit(Surv(inage, outage, status==2)~gingi.cat, data=females)
ggsurvplot(out, fun = "cumhaz", xlim=c(25,56), size=0.05, conf.int="TRUE", legend.labs=c("Gingivitis: Low", "Gingivitis: Moderate", "Gingivitis: High"), xlab ="Age during follow-up", ylab="Cumulative hazard of T2D in females")

out=survfit(Surv(inage, outage, status==2)~dmft.cat, data=males)
kmplot(out, fun="cumhaz", lty=1) 
ggsurvplot(out, fun = "cumhaz", xlim=c(25,56), size=0.05, conf.int="TRUE", legend.labs=c("DCS: Low", "DCS: Moderate", "DCS: High", "DCS: Severe"), xlab ="Age during follow-up", ylab="Cumulative hazard of T2D in males")

out=survfit(Surv(inage, outage, status==2)~gingi.cat, data=males)
ggsurvplot(out, fun = "cumhaz", xlim=c(25,56), size=0.05, conf.int="TRUE", legend.labs=c("Gingivitis: Low", "Gingivitis: Moderate", "Gingivitis: High"), xlab ="Age during follow-up", ylab="Cumulative hazard of T2D in males")


```

#Data visualisation - dmft and gingi

```{r}

#Distribution histograms 

#DMFT - population

df <- as.data.frame(table(scor$koen, scor$dmft.cat)) %>% 
  rename(Sex = Var1,
         DMFT = Var2) %>% 
  mutate(Sex = ifelse(Sex == 1, "Male", "Female"))

ggplot(df, aes (fill=Sex, x=DMFT, y=Freq))+
  geom_bar(alpha = 0.9, stat="identity", position = "dodge")+
  xlab("DC") +
  ylab("Frequency")+
  theme(plot.title = element_text(hjust=0.5), panel.background = element_blank(), axis.line = element_line(color="black"))

#DMFT - T2D
df <- scor %>% filter(status==2)
df <- as.data.frame(table(df$koen, df$dmft.cat)) %>% 
  rename(Sex = Var1,
         DMFT = Var2) %>% 
  mutate(Sex = ifelse(Sex == 1, "Male", "Female"))

ggplot(df, aes (fill=Sex, x=DMFT, y=Freq))+
  geom_bar(alpha = 0.9, stat="identity", position = "dodge")+
  xlab("DC") +
  ylab("Frequency")+
  theme(plot.title = element_text(hjust=0.5), panel.background = element_blank(), axis.line = element_line(color="black"))

#Gingi - population

df <- as.data.frame(table(scor$koen, scor$gingi.cat)) %>% 
  rename(Sex = Var1,
         Gingivitis = Var2) %>% 
  mutate(Sex = ifelse(Sex == 1, "Male", "Female"))

ggplot(df, aes (fill=Sex, x=Gingivitis, y=Freq))+
  geom_bar(alpha = 0.9, stat="identity", position = "dodge")+
  ggtitle("Gingivitis distribution by sex")+
  xlab("GS") +
  ylab("Frequency")+
  theme(plot.title = element_text(hjust=0.5), panel.background = element_blank(), axis.line = element_line(color="black"))

#Alternative
ggplot(males, aes  (x=dmft.cat))+
  geom_bar(fill = "lightblue", alpha = 0.9)+
  ggtitle("DC distribution - Females")+
  xlab("Dental caries (DMFT)") +
  ylab("Frequency")+
  theme(plot.title = element_text(hjust=0.5), panel.background = element_blank(), axis.line = element_line(color="black"))


#Jitterplot DC and gingi females:
ggplot(females, aes  (x=dmft.cat+1, y=gingi.cat+1))+
  geom_point()+
  geom_jitter()+
  scale_x_discrete(name = "DCS", limits=c("Low", "Moderate", "High", "Severe"))+
  scale_y_discrete(name = "GS", limits=c("Low", "Moderate", "High"))+
  theme_classic()

#Jitterplot DC and gingi males:
ggplot(males, aes  (x=dmft.cat+1, y=gingi.cat+1))+
  geom_point()+
  geom_jitter()+
  scale_x_discrete(name = "DCS", limits=c("Low", "Moderate", "High", "Severe"))+
  scale_y_discrete(name = "GS", limits=c("Low", "Moderate", "High"))+
  theme_classic()
    
   

qqplot(x=females$gingivitis, y=females$dmft) #non-normal distribution
qqplot(x=males$gingivitis, y=males$dmft) #non-normal distribution
